### 2023-10-06 chapters/background.tex

In society, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context. However, across most quantitative scientific fields, a consensus definition appears to align with the following \parencite{hubbard_how_2014}:
%
\begin{center}
    \textit{Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.}
\end{center}
%
Although such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability. When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain. In that sense, uncertainty is a word we use to refer to missing information, whether we know what is missing or not. 
A natural way to describe an uncertain world state is to use probabilities. To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probability theory. In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \parencite{mackay_information_2003}.

@@

\paragraph{Applications of VAEs}
Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \parencite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \parencite{pathak_context_2016}, image super-resolution \parencite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \parencite{hsu_unsupervised_2017,hsu_hierarchical_2019}. 
VAEs have also proven themselves useful for semi"=supervised learning \parencite{kingma_semi"=supervised_2014,kingma_improved_2016,maaloe_biva_2019}.

>>>>

In society, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context. However, across most quantitative scientific fields, a consensus definition appears to align with the following \parencite{hubbard_how_2014}:
%
\begin{quote}
    \centering\itshape
    Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.
\end{quote}
%
Although such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability. When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain. In that sense, uncertainty is a word we use to refer to missing information, whether we know what is missing or not. 
A natural way to describe an uncertain world state is to use probabilities. To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probability theory. In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \parencite{mackay_information_2003}.

@@

\paragraph{Applications of VAEs}
Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \parencite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \parencite{pathak_context_2016}, image super-resolution \parencite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \parencite{hsu_unsupervised_2017,hsu_hierarchical_2019}. 
VAEs have also proven themselves useful for semi"=supervised learning \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}.

### 2023-10-06 chapters/main-contributions.tex

The following chapters of the thesis are self-contained studies and therefore detail their own research hypotheses and contributions, but do so without consideration to the other chapters. 
For that reason, we here lay out research hypotheses and contributions in relation to the overall research project. 
% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self"=supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.  Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.

@@

\contribitem{part:self"=supervised-speech-representation-learning}
This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}. 
Self"=supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self"=supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}. 
In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self"=supervised learning for learning representations for unsupervised uncertainty estimation. We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}. 
Due to a desire to focus on self"=supervised methods, the paper in \cref{chp:paper-review} does not put particular focus on variational methods. 
The early version is included in \cref{app:paper-brief} for reference.

>>>>

The following chapters of the thesis are self-contained studies and therefore detail their own research hypotheses and contributions, but do so without consideration to the other chapters. 
For that reason, we here lay out research hypotheses and contributions in relation to the overall research project. 
% The chapters are structured into three parts: {\color{black}\cref{part:unsupervised-uncertainty-estimation}} is concerned with unsupervised uncertainty estimation, {\color{black}\cref{part:self-supervised-speech-representation-learning}} is concerned with self"=supervised speech representation learning, and {\color{black}\cref{part:medical-applications}} contains studies on machine learning methods applied for tasks in a medical setting.  Finally, {\color{black}\cref{part:discussion-and-conclusion}} contains a discussion and the conclusion of the thesis.

@@

\contribitem{part:self-supervised-speech-representation-learning}
This part of the thesis is concerned with self"=supervised speech representation learning and consists of a single paper that provides a substantial review of the field \parencite{mohamed_selfsupervised_2022}. 
Self"=supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self"=supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}. 
In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self"=supervised learning for learning representations for unsupervised uncertainty estimation. We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.
An early version of this review paper included a review of variational autoencoders for speech representation learning and a comparison between the two modelling paradigms \parencite{borgholt_brief_2022}. 
Due to a desire to focus on self"=supervised methods, the paper in \cref{chp:paper-review} does not put particular focus on variational methods. 
The early version is included in \cref{app:paper-brief} for reference.

### 2023-10-06 chapters/paper_benchmarking.tex

% \linepenalty=1000
% \subsection{Related work}
LVMs formulated as VAEs continue to be of interest since they are able to learn an approximation to the posterior distribution of assumed latent variables. The posterior is usually of a reduced dimensionality compared to the input and lies close to a known prior distribution. Approximate posteriors have various applications e.g. semi"=supervised learning \parencite{kingma_semi"=supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.

@@

    \begin{subtable}[t]{.45\textwidth}

@@

    \begin{adjustbox}{center, totalheight=0.35\paperheight}

@@

\paragraph{TIMIT}
For temporal resolutions of $s=1$, the deterministic autoregressive models yield the best likelihoods with WaveNet achieving $\SI{10.88}{bpf}$ on TIMIT as seen in \cref{tab: likelihoods timit}. 
Somewhat surprisingly, the LSTM baseline almost matches WaveNet with a likelihood of $\SI{11.11}{bpf}$ at $s=1$. However, due to being autoregressive in training, the LSTM trains considerably slower than the parallel convolutional WaveNet; something not conveyed by \cref{tab: likelihoods timit}. 
Notably, the VRNN and SRNN models achieve likelihoods close to that of WaveNet and the LSTM at around $\SI{11.09}{bpf}$. The STCN exhibited instability when trained at $s=1$ and tended to diverge.

@@

Decreasing the resolution $s$ improves the likelihood for all LVMs. However, the best performing models, STCN and CW-VAE are not yet scalable to this regime for reasons related to numerical instability and computational infeasibility. This seems to indicate that LVMs may be able to outperform autoregressive models at $s=1$ in the future.
\paragraph{LibriSpeech}
On LibriSpeech (\cref{tab: likelihoods librispeech}), results are similar to TIMIT. The STCN achieves the best likelihood at $s=64$ and the CW-VAE surpasses WaveNet and the LSTM. 
\paragraph{Compression}
A connection can be made between the model likelihoods and the compression rates of audio compression algorithms. 
Lossy compression algorithms, such as MP3, exploit the dynamic range of human hearing to achieve 70-95\% reduction in bit rate \parencite{brandenburg_mpeg-2_1998} while lossless compression algorithms, such as FLAC, achieve 50-70\% reduction \parencite{coalson_free_2019} independent of audio content. 
% On LibriSpeech, FLAC achieves an overall compression to about 58\% of the original bit rate.
Although both the autoregressive models and the LVMs are lossy, their objective minimizes the amount of incurred loss. 
The best likelihoods reported in tables~\cref{tab: likelihoods timit} and \cref{tab: likelihoods librispeech} correspond to about a 30\% reduction in bit rate which indicates that there are significant gains in likelihood to be made in speech modeling.

@@

\paragraph{Quantitative}
We train an ASR model to recognize phonemes and compare its performance when using input representations obtained from different unsupervised models. 
For WaveNet and the LSTM, we use the hidden state as the representation. For all LVMs, we use the latent variable. For the hierarchical LVMs and WaveNet, we run the experiment using each possible representation and report only the best one here. 
We compare the learned representations to a log Mel spectrogram with 80 filterbanks, hop length 64 and window size 128. 
We also compared to using raw PCM in vectors of 64 elements standardized to $[-1,1]$ but found that the ASR did not reliably converge at all which highlights the importance of input representation. 
The ASR model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments. 
We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit}. 
We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)}.

@@

\section*{Acknowledgements}
This research was partially funded by the Innovation Fund Denmark via the Industrial PhD Programme (grant no.\@ 0153-00167B). JF and SH were funded in part by the Novo Nordisk Foundation (grant no.\@ NNF20OC0062606) via the Center for Basic Machine Learning Research in Life Science (MLLS, \hyperlink{https://www.mlls.dk}{https://www.mlls.dk}). JF was further funded by the Novo Nordisk Foundation (grant no.\@ NNF20OC0065611) and the Independent Research Fund Denmark (grant no.\@ 9131-00082B). SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement no. 757360).

>>>>

% \linepenalty=1000
% \subsection{Related work}
LVMs formulated as VAEs continue to be of interest since they are able to learn an approximation to the posterior distribution of assumed latent variables. The posterior is usually of a reduced dimensionality compared to the input and lies close to a known prior distribution. Approximate posteriors have various applications e.g. semi"=supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.

@@

    \begin{subtable}[t]{.49\textwidth}

@@

    \begin{adjustbox}{center, totalheight=0.4\paperheight}

@@

\paragraph{TIMIT}
For temporal resolutions of $s=1$, the deterministic autoregressive models yield the best likelihoods with WaveNet achieving $\SI{10.88}{bpf}$ on TIMIT as seen in \cref{tab: likelihoods timit} (left). 
Somewhat surprisingly, the LSTM baseline almost matches WaveNet with a likelihood of $\SI{11.11}{bpf}$ at $s=1$. However, due to being autoregressive in training, the LSTM trains considerably slower than the parallel convolutional WaveNet; something not conveyed by \cref{tab: likelihoods timit} (left). 
Notably, the VRNN and SRNN models achieve likelihoods close to that of WaveNet and the LSTM at around $\SI{11.09}{bpf}$. The STCN exhibited instability when trained at $s=1$ and tended to diverge.

@@

Decreasing the resolution $s$ improves the likelihood for all LVMs. However, the best performing models, STCN and CW-VAE are not yet scalable to this regime for reasons related to numerical instability and computational infeasibility. This seems to indicate that LVMs may be able to outperform autoregressive models at $s=1$ in the future.
\paragraph{LibriSpeech}
On LibriSpeech (\cref{tab: likelihoods librispeech}), results are similar to TIMIT. The STCN achieves the best likelihood at $s=64$ and the CW-VAE surpasses WaveNet and the LSTM. 
\paragraph{Compression}
A connection can be made between the model likelihoods and the compression rates of audio compression algorithms. 
Lossy compression algorithms, such as MP3, exploit the dynamic range of human hearing to achieve 70-95\% reduction in bit rate \parencite{brandenburg_mpeg-2_1998} while lossless compression algorithms, such as FLAC, achieve 50-70\% reduction \parencite{coalson_free_2019} independent of audio content. 
% On LibriSpeech, FLAC achieves an overall compression to about 58\% of the original bit rate.
Although both the autoregressive models and the LVMs are lossy, their objective minimizes the amount of incurred loss. 
The best likelihoods reported in \cref{tab: likelihoods timit} (left) and \cref{tab: likelihoods librispeech} correspond to about a 30\% reduction in bit rate which indicates that there are significant gains in likelihood to be made in speech modeling.

@@

\paragraph{Quantitative}
We train an ASR model to recognize phonemes and compare its performance when using input representations obtained from different unsupervised models. 
For WaveNet and the LSTM, we use the hidden state as the representation. For all LVMs, we use the latent variable. For the hierarchical LVMs and WaveNet, we run the experiment using each possible representation and report only the best one here. 
We compare the learned representations to a log Mel spectrogram with 80 filterbanks, hop length 64 and window size 128. 
We also compared to using raw PCM in vectors of 64 elements standardized to $[-1,1]$ but found that the ASR did not reliably converge at all which highlights the importance of input representation. 
The ASR model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments. 
We pre-train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).

@@

\section*{Acknowledgements}
This research was partially funded by the Innovation Fund Denmark via the Industrial PhD Programme (grant no.\@ 0153-00167B). JF and SH were funded in part by the Novo Nordisk Foundation (grant no.\@ NNF20OC0062606) via the Center for Basic Machine Learning Research in Life Science (MLLS, \hyperlink{https://www.mlls.dk}{https://www.mlls.dk}). JF was further funded by the Novo Nordisk Foundation (grant no.\@ NNF20OC0065611) and the Independent Research Fund Denmark (grant no.\@ 9131-00082B). SH was further funded by VILLUM FONDEN (15334) and the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement no. 757360).
